Endogenous Macrodynamics in Algorithmic Recourse

### 2023-01-03 paper/paper.tex
grammarly

Among the approaches to XAI that have recently grown in popularity are \textbf{Counterfactual Explanations} (CE). They explain how inputs into a model need to change for it to produce different outputs. Counterfactual Explanations that involve realistic and actionable changes can be used for the purpose of \textbf{Algorithmic Recourse} (AR) to help individuals who face adverse outcomes. An example relevant to the Social Sciences is consumer credit: in this context, AR can be used to guide individuals in improving their creditworthiness, should they have previously been denied access to credit based on an automated decision-making system. A meaningful recourse recommendation for a denied applicant could be: \emph{``If your net savings rate had been 10\% of your monthly income instead of the actual 8\%, your application would have been successful. See if you can temporarily cut down on consumption.''} In the remainder of this paper we will use both terminologies---recourse and counterfactual---interchangeably to refer to situations where counterfactuals are generated with the intent to provide individual recourse.

@@

\begin{example}[Consumer Credit]
\protect\hypertarget{exm:consumer}{}\label{exm:consumer}Suppose Figure \ref{fig:poc} relates to an automated decision-making system used by a retail bank to evaluate credit applicants with respect to their creditworthiness. Assume that the two features are actually meaningful in the sense that creditworthiness increases in the south-east direction. Then we can think of the outcome in panel (d) as representing a situation where the bank supplies credit to more borrowers (orange), but these borrowers are on average less creditworthy and more of them can be expected to default on their loan. This represents a cost to the retail bank.
\end{example}

\begin{example}[Student Admission]
\protect\hypertarget{exm:student}{}\label{exm:student}Suppose Figure \ref{fig:poc} relates to an automated decision-making system used by a university in their student admission process. Assume that the two features are actually meaningful in the sense that the likelihood of students successfully completing their degree increases in the south-east direction. Then we can think of the outcome in panel (b) as representing a situation where more students are admitted to university (orange), but they are more likely to fail their degree than students that were admitted in previous years. The university admission committee catches on to this and suspends its efforts to offer Algorithmic Recourse. This represents an opportunity cost to future student applicants, that may have derived utility from being offered recourse.
\end{example}

@@

To the best of our knowledge this is the first work investigating endogenous macrodynamics in AR. Our contributions to the state of knowledge are as follows: firstly, we posit a compelling argument that calls for a novel perspective on Algorithmic Recourse extending our focus from single individuals to groups (Sections \ref{related} and \ref{method}). Secondly, we introduce an experimental framework extending previous work by Altmeyer \protect\hyperlink{ref-altmeyer2022counterfactualexplanations}{{[}5{]}}, which enables us to study macrodynamics of Algorithmic Recourse through simulations that can be fully parallelized (Section \ref{method-2}). Thirdly, we use this framework to provide a first in-depth analysis of endogenous recourse dynamics induced by various popular counterfactual generators including \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}}, \protect\hyperlink{ref-schut2021generating}{{[}6{]}}, \protect\hyperlink{ref-joshi2019realistic}{{[}7{]}}, \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} and \protect\hyperlink{ref-antoran2020getting}{{[}9{]}} (Sections \ref{empirical} and \ref{empirical-2}). Fourthly, given that we find substantial impact of recourse, we propose key mitigation strategies and measure their impact experimentally (Section \ref{mitigate}). Finally, we discuss our findings in the broader context of the literature in Section \ref{discussion}, before pointing to some of the limitations of our work as well as avenues for future research in Section \ref{limit}. Section \ref{conclusion} concludes.

@@

In this section we provide a review of the relevant literature. First, Subsection \ref{related-recourse} discusses the existing research within the domain of Counterfactual Explanations and Algorithmic Recourse. Then, Subsection \ref{related-shifts} presents some of the previous work on the measurement of dataset and model shifts.

@@

As mentioned earlier, research has so far also generally focused on generating counterfactuals for single individuals or instances. We have been able to identify only one existing work that investigates black-box model behavior towards a group of individuals \protect\hyperlink{ref-carrizosa2021generating}{{[}15{]}}. The authors propose an optimization framework that generates collective counterfactuals. We provide a motivation for doing so from the perspective of endogenous macrodynamics of Algorithmic Recourse.

@@

Much attention has been paid to the detection of dataset shifts -- situations where the distribution of data changes over time. Rabanser et al.~suggest a framework to detect data drift from a minimal number of samples through the application of two-sample tests \protect\hyperlink{ref-rabanser2019failing}{{[}16{]}}. This task is a generalization of the anomaly detection problem for large datasets, which aims to answer the question if two sets of samples could have been generated from the same probability distribution. Numerous approaches to anomaly detection have been summarized \protect\hyperlink{ref-chandola2009anomaly}{{[}17{]}}. Another well-established research topic is that of concept drift: situations where external variables influence the patterns between the input and the output of a model \protect\hyperlink{ref-widmer1996learning}{{[}18{]}}. For instance, Gama et al.~offer a review of the adaptive learning techniques which can handle concept drift \protect\hyperlink{ref-gama2014survey}{{[}19{]}}. Less previous work is available on the related topic of model drift: changes in model performance over time. Nelson et al.~review how resistant different machine learning models are to the model drift \protect\hyperlink{ref-nelson2015evaluating}{{[}20{]}}. Ackerman et al.~offer a method to detect changes in model performance when ground truth is not available \protect\hyperlink{ref-ackerman2021machine}{{[}21{]}}.

In the context of Algorithmic Recourse, domain and model shifts were first brought up by the authors behind ROAR \protect\hyperlink{ref-upadhyay2021robust}{{[}14{]}}. In their work they refer to model shifts as simply any perturbation \(\Delta\) to the parameters of the model in question: \(M\). While this also sets the baseline for our analysis here, it is worth noting that in \protect\hyperlink{ref-upadhyay2021robust}{{[}14{]}} these perturbations are mechanically introduced. In contrast, we are interested in quantifying model shifts that arise endogenously as part of a dynamic recourse process. In addition to quantifying the magnitude of shifts \(\Delta\), we aim to also analyse the characteristics of changes to the model, such as the position of the decision boundary and the overall decisiveness of the model. We have not been able to identify previous work on this topic.

@@

In this section we first set out a generalized framework for gradient-based counterfactual search that encapsulates the various individual recourse methods we have chosen to use in our experiments (Section \ref{method-general}). We then introduce the notion of a hidden external cost in algorithmic recourse and extend the existing framework to explicitly address this cost in the counterfactual search objective (Section \ref{method-collective}).

@@

We have chosen to focus on gradient-based counterfactual search for two reasons: firstly, they can be seen as direct descendants of our baseline method (Wachter); secondly, gradient-based search is particularly well-suited for differentiable black-box models like deep neural networks, which we focus on in this work. In particular, we include the following generators in our simulation experiments below: \textbf{REVISE} \protect\hyperlink{ref-joshi2019realistic}{{[}7{]}}, \textbf{CLUE} \protect\hyperlink{ref-antoran2020getting}{{[}9{]}}, \textbf{DiCE} \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} and a greedy approach that relies on probabilistic models \protect\hyperlink{ref-schut2021generating}{{[}6{]}}. Our motivation for including these different generators in our analysis, is that they all offer slightly different approaches to generate meaningful counterfactuals for differentiable black-box models. We hypothesize that generating more \textbf{meaningful} counterfactuals should mitigate the endogenous dynamics illustrated in Figure \ref{fig:poc} in Section \ref{intro}. This intuition stems from the underlying idea that more meaningful counterfactuals are generated by the same or at least a very similar data generating process as the training data. All else equal, counterfactuals that fulfill this basic requirement should be less prone to trigger shifts.

@@

Here \(\mathbf{s}^\prime=\left\{s_k^\prime\right\}_K\) is a \(K\)-dimensional array of counterfactual states and \(f: \mathcal{S} \mapsto \mathcal{X}\) maps from the counterfactual state space to the feature space. In Wachter, the state space is the feature space: \(f\) is the identity function and the number of counterfactuals \(K\) is one. Both REVISE and CLUE search counterfactuals in some latent space \(\mathcal{S}\) instead of the feature space directly. The latent embedding is learned by a separate generative model that is tasked with learning the data generating process (DGP) of \(X\). In this case, \(f\) in Equation \eqref{eq:general} corresponds to the decoder part of the generative model, that is the function that maps back from the latent space to inputs. Provided the generative model is well-specified, traversing the latent embedding typically yields meaningful counterfactuals, since they are implicitly generated by the (learned) DGP \protect\hyperlink{ref-joshi2019realistic}{{[}7{]}}.

CLUE distinguishes itself from REVISE and other counterfactual generators in that it aims to minimize the predictive uncertainty of the model in question, \(M\). To quantify predictive uncertainty, Antoran et al. \protect\hyperlink{ref-antoran2020getting}{{[}9{]}} rely on entropy estimates for probabilistic models. The greedy approach proposed by Schut et al. \protect\hyperlink{ref-schut2021generating}{{[}6{]}}, which we refer to as \textbf{Greedy}, also works with the subclass of models \(\tilde{\mathcal{M}}\subset\mathcal{M}\) that can produce predictive uncertainty estimates. The authors show that in this setting the cost function \(\text{cost}(\cdot)\) in Equation \eqref{eq:general} is redundant and meaningful counterfactuals can be generated in a fast and efficient manner through a modified Jacobian-based Saliency Map Attack (JSMA). Schut et al. \protect\hyperlink{ref-schut2021generating}{{[}6{]}} also show that by maximizing the predicted probability of \(x^\prime\) being assigned to target class \(y^*\), we also implicitly minimize predictive entropy (as in CLUE). In that sense, CLUE can be seen as equivalent to REVISE in the Bayesian context and we shall therefore refer to both approaches collectively as \textbf{Latent Space} generators\footnote{In fact, there are a number of other recently proposed approaches to counterfactual search that also broadly fall in this same category. They largely differ with respect to the chosen generative model: for example, the generator proposed by Dombrowski et al. \protect\hyperlink{ref-dombrowski2021diffeomorphic}{{[}23{]}} relies on normalizing flows.}.

Finally, DiCE \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} distinguishes itself from all other generators considered here in that it aims to generate a diverse set of \(K>1\) counterfactuals. Wachter et al. \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}} show that diverse outcomes can in principal be achieved simply rerunning counterfactual search multiple times using stochastic gradient descent (or by randomly initializing the counterfactual)\footnote{Note, in fact, that \eqref{eq:general} naturally lends itself to that idea: setting \(K\) to some value greater than one and using the Wachter objective essentially boils down to computing multiple counterfactuals in parallel. Here, \(yloss(\cdot)\) is first broadcasted over elements of \(\mathbf{s}^\prime\) and then aggregated. This is exactly how counterfactual search is implemented in \href{https://github.com/pat-alt/CounterfactualExplanations.jl}{\texttt{CounterfactualExplanations.jl}}.}. In \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} diversity is explicitly proxied via Determinantal Point Processes (DDP): the authors simply introduce DDP as a component of the cost function \(\text{cost}(\mathbf{s}^\prime)\) and thereby produce counterfactuals \(s_1, ... , s_K\) that look as different from each other as possible. The implementation of DiCE in our library of choice---\href{https://github.com/pat-alt/CounterfactualExplanations.jl}{\texttt{CounterfactualExplanations.jl}}---uses that exact approach. It is worth noting that for \(k=1\), DiCE reduces to Wachter since the DDP is constant and therefore does not affect the objective function in Equation \eqref{eq:general}.

@@

Here \(\text{cost}(f(\mathbf{s}^\prime))\) denotes the proxy for private costs faced by the individual as before and \(\lambda_1\) governs to what extent that private cost ought to be penalized. The newly introduced term \(\text{extcost}(f(\mathbf{s}^\prime))\) is meant to capture and address external costs incurred by the collective of individuals in response to changes in \(\mathbf{s}^\prime\). The underlying concept of private and external costs is borrowed from Economics and well-established in that field: when the decisions or actions by some individual market participant generate external costs, then the market is said to suffer from negative externalities and considered inefficient \protect\hyperlink{ref-pindyck2014microeconomics}{{[}24{]}}. We think that this concept describes the endogenous dynamics of algorithmic recourse observed here very well. As with individual recourse, the exact choice of \(\text{extcost}(\cdot)\) is not obvious, nor do we intend to provide a definite answer in this work, if such even exists. That being said, we do propose a few potential mitigation strategies in Section \ref{mitigate}.

@@

In the following we describe the framework we propose for modeling and analyzing endogenous macrodynamics in Algorithmic Recourse. We introduce this framework with the ambition to shed light on the following research questions:

@@

\begin{proposition}[Heterogeneity]
\protect\hypertarget{prp:het}{}\label{prp:het}Do different counterfactual generators yield significantly different outcomes in this context? Furthermore, is there any heterogeneity with respect to the chosen classifier and dataset?
\end{proposition}

\begin{proposition}[Drivers]
\protect\hypertarget{prp:drive}{}\label{prp:drive}What are drivers of endogenous dynamics in Algorithmic Recourse?
\end{proposition}

@@

The dynamics illustrated in Figure \ref{fig:poc} in were generated through a simple experiment that aims to simulate the process of Algorithmic Recourse in practice. We begin in the static setting at time time \(t=0\): firstly, we have some binary classifier \(M\) that was pre-trained on data \(\mathcal{D}=\mathcal{D}_0 \cup \mathcal{D}_1\), where \(\mathcal{D}_0\) and \(\mathcal{D}_1\) denote samples in the non-target and target class, respectively; secondly, we generate recourse for a random batch of \(B\) individuals in the non-target class (\(\mathcal{D}_0\)). Note that we focus our attention on classification problems, since classification poses the most common use-case for recourse\footnote{To keep notation simple, we have also restricted ourselves to binary classification here, but \texttt{AlgorithmicRecourseDynamics.jl} can also be used for multi-class problems.}.

In order to simulate the dynamic process, we suppose that the model \(M\) is retrained following the actual implementation of recourse in time \(t=0\). Following the update to the model, we assume that at time \(t=1\) recourse is generated for yet another random subset of individuals in the non-target class. This process is repeated for a number of time periods \(T\). To get a clean read on endogenous dynamics we keep the total population of samples closed: we allow existing samples to move from factual to counterfactual states, but do not allow any entirely new samples to enter the population. The experimental setup is summarized in Algorithm \ref{algo-experiment}.

@@

Along with any other fixed parameters affecting the counterfactual search, the parameters \(T\) and \(B\) are assumed as given in Algorithm \ref{algo-experiment}. Still, it worth noting that the higher these values, the more factual instances undergo recourse throughout the entire experiment. Of course, this is likely to lead to more pronounced domain and model shifts by time \(T\). In our experiments, we choose the values such that \(T \cdot B\) corresponds to the application of recourse on \(\approx50\%\) of the negative instances from the initial dataset. As we compute evaluation metrics at regular intervals throughout the procedure, we can also verify the impact of recourse when it is implemented for a smaller number of individuals.

Algorithm \ref{algo-experiment} summarizes the proposed simulation experiment for a given dataset \(\mathcal{D}\), model \(M\) and generator \(G\), but naturally we are interested in comparing simulation outcomes for different sources of data, models and generators. The framework we have built facilitates this, making use of multi-threading in order to speed up computations. Holding the initial model and dataset constant, the experiments are run for all generators, since our primary concern is to benchmark different recourse methods. To ensure that each generator is faced with the same initial conditions in each round \(t\), the candidate batch of individuals from the non-target class is randomly drawn from the intersection of all non-target class individuals across all experiments \(\left\{\textsc{Experiment}(M,\mathcal{D},G)\right\}_{j=1}^J\) where \(J\) is the total number of generators.

@@

We formulate two desiderata for the set of metrics used to measure domain and model shifts induced by recourse. First, the metrics should be applicable regardless of the dataset or classification technique so that they allow for the meaningful comparison of the generators in various scenarios. As the knowledge of the underlying probability distribution is rarely available, the metrics should be empirical and non-parametric. This further ensures that we can also measure large datasets by sampling from the available data. Moreover, while our study was conducted in a two-class classification setting, our choice of metrics should remain applicable in the future research on multi- class recourse problems. Second, the set of metrics should allow to capture various aspects of the previously mentioned magnitude, path, and tempo of changes while remaining as small as possible.

@@

As our baseline for quantifying model shifts we measure perturbations to the model parameters at each point in time \(t\) following \protect\hyperlink{ref-upadhyay2021robust}{{[}14{]}}. We define \(\Delta=||\theta_{t+1}-\theta_{t}||^2\), that is the euclidean distance between the vectors of parameters before and after retraining the model \(M\). We shall refer to this baseline metric simply as \textbf{Perturbations}.

We extend the metric in Equation \eqref{eq:mmd} for the purpose of quantifying model shifts. Specifically, we introduce \textbf{Predicted Probability MMD (PP MMD)}: instead of applying Equation \eqref{eq:mmd} to features directly, we apply it to the predicted probabilities assigned to a set of samples by the model \(M\). If the model shifts, the probabilities assigned to each sample will change; again, this metric will equal 0 only if the two classifiers are the same. We compute PP MMD in two ways: firstly, we compute it over samples drawn uniformly from the dataset, and, secondly, we compute it over points spanning a mesh grid over a subspace of the entire feature space. For the latter approach we bound the subspace by the extrema of each feature. While this approach is theoretically more robust, unfortunately, it suffers from the curse of dimensionality, since it becomes increasingly difficult to select enough points to overcome noise as the dimension \(D\) grows.

>>>>

Among the approaches to XAI that have recently grown in popularity are \textbf{Counterfactual Explanations} (CE). They explain how inputs into a model need to change for it to produce different outputs. Counterfactual Explanations that involve realistic and actionable changes can be used for the purpose of \textbf{Algorithmic Recourse} (AR) to help individuals who face adverse outcomes. An example relevant to the Social Sciences is consumer credit: in this context, AR can be used to guide individuals in improving their creditworthiness, should they have previously been denied access to credit based on an automated decision-making system. A meaningful recourse recommendation for a denied applicant could be: \emph{``If your net savings rate had been 10\% of your monthly income instead of the actual 8\%, your application would have been successful. See if you can temporarily cut down on consumption.''} In the remainder of this paper, we will use both terminologies---recourse and counterfactual---interchangeably to refer to situations where counterfactuals are generated with the intent to provide individual recourse.

@@

\begin{example}[Consumer Credit]
\protect\hypertarget{exm:consumer}{}\label{exm:consumer}Suppose Figure \ref{fig:poc} relates to an automated decision-making system used by a retail bank to evaluate credit applicants with respect to their creditworthiness. Assume that the two features are meaningful in the sense that creditworthiness increases in the south-east direction. Then we can think of the outcome in panel (d) as representing a situation where the bank supplies credit to more borrowers (orange), but these borrowers are on average less creditworthy and more of them can be expected to default on their loan. This represents a cost to the retail bank.
\end{example}

\begin{example}[Student Admission]
\protect\hypertarget{exm:student}{}\label{exm:student}Suppose Figure \ref{fig:poc} relates to an automated decision-making system used by a university in its student admission process. Assume that the two features are meaningful in the sense that the likelihood of students completing their degree increases in the south-east direction. Then we can think of the outcome in panel (b) as representing a situation where more students are admitted to university (orange), but they are more likely to fail their degree than students that were admitted in previous years. The university admission committee catches on to this and suspends its efforts to offer Algorithmic Recourse. This represents an opportunity cost to future student applicants, that may have derived utility from being offered recourse.
\end{example}

@@

To the best of our knowledge, this is the first work investigating endogenous macrodynamics in AR. Our contributions to the state of knowledge are as follows: firstly, we posit a compelling argument that calls for a novel perspective on Algorithmic Recourse extending our focus from single individuals to groups (Sections \ref{related} and \ref{method}). Secondly, we introduce an experimental framework extending previous work by Altmeyer \protect\hyperlink{ref-altmeyer2022counterfactualexplanations}{{[}5{]}}, which enables us to study macrodynamics of Algorithmic Recourse through simulations that can be fully parallelized (Section \ref{method-2}). Thirdly, we use this framework to provide a first in-depth analysis of endogenous recourse dynamics induced by various popular counterfactual generators including \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}}, \protect\hyperlink{ref-schut2021generating}{{[}6{]}}, \protect\hyperlink{ref-joshi2019realistic}{{[}7{]}}, \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} and \protect\hyperlink{ref-antoran2020getting}{{[}9{]}} (Sections \ref{empirical} and \ref{empirical-2}). Fourthly, given that we find a substantial impact of recourse, we propose key mitigation strategies and measure their impact experimentally (Section \ref{mitigate}). Finally, we discuss our findings in the broader context of the literature in Section \ref{discussion}, before pointing to some of the limitations of our work as well as avenues for future research in Section \ref{limit}. Section \ref{conclusion} concludes.

@@

In this section, we provide a review of the relevant literature. First, Subsection \ref{related-recourse} discusses the existing research within the domain of Counterfactual Explanations and Algorithmic Recourse. Then, Subsection \ref{related-shifts} presents some of the previous work on the measurement of dataset and model shifts.

@@

As mentioned earlier, research has so far also generally focused on generating counterfactuals for single individuals or instances. We have been able to identify only one existing work that investigates black-box model behaviour towards a group of individuals \protect\hyperlink{ref-carrizosa2021generating}{{[}15{]}}. The authors propose an optimization framework that generates collective counterfactuals. We provide a motivation for doing so from the perspective of endogenous macrodynamics of Algorithmic Recourse.

@@

Much attention has been paid to the detection of dataset shifts -- situations where the distribution of data changes over time. Rabanser et al.~suggest a framework to detect data drift from a minimal number of samples through the application of two-sample tests \protect\hyperlink{ref-rabanser2019failing}{{[}16{]}}. This task is a generalization of the anomaly detection problem for large datasets, which aims to answer the question if two sets of samples could have been generated from the same probability distribution. Numerous approaches to anomaly detection have been summarized \protect\hyperlink{ref-chandola2009anomaly}{{[}17{]}}. Another well-established research topic is concept drift: situations where external variables influence the patterns between the input and the output of a model \protect\hyperlink{ref-widmer1996learning}{{[}18{]}}. For instance, Gama et al.~offer a review of the adaptive learning techniques which can handle concept drift \protect\hyperlink{ref-gama2014survey}{{[}19{]}}. Less previous work is available on the related topic of model drift: changes in model performance over time. Nelson et al.~review how resistant different machine learning models are to model drift \protect\hyperlink{ref-nelson2015evaluating}{{[}20{]}}. Ackerman et al.~offer a method to detect changes in model performance when ground truth is not available \protect\hyperlink{ref-ackerman2021machine}{{[}21{]}}.

In the context of Algorithmic Recourse, domain and model shifts were first brought up by the authors behind ROAR \protect\hyperlink{ref-upadhyay2021robust}{{[}14{]}}. In their work, they refer to model shifts as simply any perturbation \(\Delta\) to the parameters of the model in question: \(M\). While this also sets the baseline for our analysis here, it is worth noting that in \protect\hyperlink{ref-upadhyay2021robust}{{[}14{]}} these perturbations are mechanically introduced. In contrast, we are interested in quantifying model shifts that arise endogenously as part of a dynamic recourse process. In addition to quantifying the magnitude of shifts \(\Delta\), we aim to also analyse the characteristics of changes to the model, such as the position of the decision boundary and the overall decisiveness of the model. We have not been able to identify previous work on this topic.

@@

In this section, we first set out a generalized framework for gradient-based counterfactual search that encapsulates the various individual recourse methods we have chosen to use in our experiments (Section \ref{method-general}). We then introduce the notion of a hidden external cost in algorithmic recourse and extend the existing framework to explicitly address this cost in the counterfactual search objective (Section \ref{method-collective}).

@@

We have chosen to focus on gradient-based counterfactual search for two reasons: firstly, they can be seen as direct descendants of our baseline method (Wachter); secondly, gradient-based search is particularly well-suited for differentiable black-box models like deep neural networks, which we focus on in this work. In particular, we include the following generators in our simulation experiments below: \textbf{REVISE} \protect\hyperlink{ref-joshi2019realistic}{{[}7{]}}, \textbf{CLUE} \protect\hyperlink{ref-antoran2020getting}{{[}9{]}}, \textbf{DiCE} \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} and a greedy approach that relies on probabilistic models \protect\hyperlink{ref-schut2021generating}{{[}6{]}}. Our motivation for including these different generators in our analysis is that they all offer slightly different approaches to generating meaningful counterfactuals for differentiable black-box models. We hypothesize that generating more \textbf{meaningful} counterfactuals should mitigate the endogenous dynamics illustrated in Figure \ref{fig:poc} in Section \ref{intro}. This intuition stems from the underlying idea that more meaningful counterfactuals are generated by the same or at least a very similar data-generating process as the training data. All else equal, counterfactuals that fulfil this basic requirement should be less prone to trigger shifts.

@@

Here \(\mathbf{s}^\prime=\left\{s_k^\prime\right\}_K\) is a \(K\)-dimensional array of counterfactual states and \(f: \mathcal{S} \mapsto \mathcal{X}\) maps from the counterfactual state space to the feature space. In Wachter, the state space is the feature space: \(f\) is the identity function and the number of counterfactuals \(K\) is one. Both REVISE and CLUE search counterfactuals in some latent space \(\mathcal{S}\) instead of the feature space directly. The latent embedding is learned by a separate generative model that is tasked with learning the data-generating process (DGP) of \(X\). In this case, \(f\) in Equation \eqref{eq:general} corresponds to the decoder part of the generative model, that is the function that maps back from the latent space to inputs. Provided the generative model is well-specified, traversing the latent embedding typically yields meaningful counterfactuals since they are implicitly generated by the (learned) DGP \protect\hyperlink{ref-joshi2019realistic}{{[}7{]}}.

CLUE distinguishes itself from REVISE and other counterfactual generators in that it aims to minimize the predictive uncertainty of the model in question, \(M\). To quantify predictive uncertainty, Antoran et al. \protect\hyperlink{ref-antoran2020getting}{{[}9{]}} rely on entropy estimates for probabilistic models. The greedy approach proposed by Schut et al. \protect\hyperlink{ref-schut2021generating}{{[}6{]}}, which we refer to as \textbf{Greedy}, also works with the subclass of models \(\tilde{\mathcal{M}}\subset\mathcal{M}\) that can produce predictive uncertainty estimates. The authors show that in this setting the cost function \(\text{cost}(\cdot)\) in Equation \eqref{eq:general} is redundant and meaningful counterfactuals can be generated in a fast and efficient manner through a modified Jacobian-based Saliency Map Attack (JSMA). Schut et al. \protect\hyperlink{ref-schut2021generating}{{[}6{]}} also show that by maximizing the predicted probability of \(x^\prime\) being assigned to target class \(y^*\), we also implicitly minimize predictive entropy (as in CLUE). In that sense, CLUE can be seen as equivalent to REVISE in the Bayesian context and we shall therefore refer to both approaches collectively as \textbf{Latent Space} generators\footnote{In fact, there are several other recently proposed approaches to counterfactual search that also broadly fall in this same category. They largely differ with respect to the chosen generative model: for example, the generator proposed by Dombrowski et al. \protect\hyperlink{ref-dombrowski2021diffeomorphic}{{[}23{]}} relies on normalizing flows.}.

Finally, DiCE \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} distinguishes itself from all other generators considered here in that it aims to generate a diverse set of \(K>1\) counterfactuals. Wachter et al. \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}} show that diverse outcomes can in principle be achieved simply by rerunning counterfactual search multiple times using stochastic gradient descent (or by randomly initializing the counterfactual)\footnote{Note that \eqref{eq:general} naturally lends itself to that idea: setting \(K\) to some value greater than one and using the Wachter objective essentially boils down to computing multiple counterfactuals in parallel. Here, \(yloss(\cdot)\) is first broadcasted over elements of \(\mathbf{s}^\prime\) and then aggregated. This is exactly how counterfactual search is implemented in \href{https://github.com/pat-alt/CounterfactualExplanations.jl}{\texttt{CounterfactualExplanations.jl}}.}. In \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} diversity is explicitly proxied via Determinantal Point Processes (DDP): the authors simply introduce DDP as a component of the cost function \(\text{cost}(\mathbf{s}^\prime)\) and thereby produce counterfactuals \(s_1, ..., s_K\) that look as different from each other as possible. The implementation of DiCE in our library of choice---\href{https://github.com/pat-alt/CounterfactualExplanations.jl}{\texttt{CounterfactualExplanations.jl}}---uses that exact approach. It is worth noting that for \(k=1\), DiCE reduces to Wachter since the DDP is constant and therefore does not affect the objective function in Equation \eqref{eq:general}.

@@

Here \(\text{cost}(f(\mathbf{s}^\prime))\) denotes the proxy for private costs faced by the individual as before and \(\lambda_1\) governs to what extent that private cost ought to be penalized. The newly introduced term \(\text{extcost}(f(\mathbf{s}^\prime))\) is meant to capture and address external costs incurred by the collective of individuals in response to changes in \(\mathbf{s}^\prime\). The underlying concept of private and external costs is borrowed from Economics and well-established in that field: when the decisions or actions by some individual market participant generate external costs, then the market is said to suffer from negative externalities and is considered inefficient \protect\hyperlink{ref-pindyck2014microeconomics}{{[}24{]}}. We think that this concept describes the endogenous dynamics of algorithmic recourse observed here very well. As with individual recourse, the exact choice of \(\text{extcost}(\cdot)\) is not obvious, nor do we intend to provide a definite answer in this work, if such even exists. That being said, we do propose a few potential mitigation strategies in Section \ref{mitigate}.

@@

In the following, we describe the framework we propose for modelling and analyzing endogenous macrodynamics in Algorithmic Recourse. We introduce this framework with the ambition to shed light on the following research questions:

@@

\begin{proposition}[Heterogeneity]
\protect\hypertarget{prp:het}{}\label{prp:het}Do different counterfactual generators yield significantly different outcomes in this context? Furthermore, is there any heterogeneity concerning the chosen classifier and dataset?
\end{proposition}

\begin{proposition}[Drivers]
\protect\hypertarget{prp:drive}{}\label{prp:drive}What are the drivers of endogenous dynamics in Algorithmic Recourse?
\end{proposition}

@@

The dynamics illustrated in Figure \ref{fig:poc} were generated through a simple experiment that aims to simulate the process of Algorithmic Recourse in practice. We begin in the static setting at time \(t=0\): firstly, we have some binary classifier \(M\) that was pre-trained on data \(\mathcal{D}=\mathcal{D}_0 \cup \mathcal{D}_1\), where \(\mathcal{D}_0\) and \(\mathcal{D}_1\) denote samples in the non-target and target class, respectively; secondly, we generate recourse for a random batch of \(B\) individuals in the non-target class (\(\mathcal{D}_0\)). Note that we focus our attention on classification problems since classification poses the most common use-case for recourse\footnote{To keep notation simple, we have also restricted ourselves to binary classification here, but \texttt{AlgorithmicRecourseDynamics.jl} can also be used for multi-class problems.}.

In order to simulate the dynamic process, we suppose that the model \(M\) is retrained following the actual implementation of recourse in time \(t=0\). Following the update to the model, we assume that at time \(t=1\) recourse is generated for yet another random subset of individuals in the non-target class. This process is repeated for a number of time periods \(T\). To get a clean read on endogenous dynamics we keep the total population of samples closed: we allow existing samples to move from factual to counterfactual states but do not allow any entirely new samples to enter the population. The experimental setup is summarized in Algorithm \ref{algo-experiment}.

@@

Along with any other fixed parameters affecting the counterfactual search, the parameters \(T\) and \(B\) are assumed as given in Algorithm \ref{algo-experiment}. Still, it is worth noting that the higher these values, the more factual instances undergo recourse throughout the entire experiment. Of course, this is likely to lead to more pronounced domain and model shifts by time \(T\). In our experiments, we choose the values such that \(T \cdot B\) corresponds to the application of recourse on \(\approx50\%\) of the negative instances from the initial dataset. As we compute evaluation metrics at regular intervals throughout the procedure, we can also verify the impact of recourse when it is implemented for a smaller number of individuals.

Algorithm \ref{algo-experiment} summarizes the proposed simulation experiment for a given dataset \(\mathcal{D}\), model \(M\) and generator \(G\), but naturally, we are interested in comparing simulation outcomes for different sources of data, models and generators. The framework we have built facilitates this, making use of multi-threading in order to speed up computations. Holding the initial model and dataset constant, the experiments are run for all generators, since our primary concern is to benchmark different recourse methods. To ensure that each generator is faced with the same initial conditions in each round \(t\), the candidate batch of individuals from the non-target class is randomly drawn from the intersection of all non-target class individuals across all experiments \(\left\{\textsc{Experiment}(M,\mathcal{D},G)\right\}_{j=1}^J\) where \(J\) is the total number of generators.

@@

We formulate two desiderata for the set of metrics used to measure domain and model shifts induced by recourse. First, the metrics should be applicable regardless of the dataset or classification technique so that they allow for the meaningful comparison of the generators in various scenarios. As knowledge of the underlying probability distribution is rarely available, the metrics should be empirical and non-parametric. This further ensures that we can also measure large datasets by sampling from the available data. Moreover, while our study was conducted in a two-class classification setting, our choice of metrics should remain applicable in future research on multi-class recourse problems. Second, the set of metrics should allow capturing various aspects of the previously mentioned magnitude, path, and tempo of changes while remaining as small as possible.

@@

As our baseline for quantifying model shifts, we measure perturbations to the model parameters at each point in time \(t\) following \protect\hyperlink{ref-upadhyay2021robust}{{[}14{]}}. We define \(\Delta=||\theta_{t+1}-\theta_{t}||^2\), that is the euclidean distance between the vectors of parameters before and after retraining the model \(M\). We shall refer to this baseline metric simply as \textbf{Perturbations}.

We extend the metric in Equation \eqref{eq:mmd} to quantify model shifts. Specifically, we introduce \textbf{Predicted Probability MMD (PP MMD)}: instead of applying Equation \eqref{eq:mmd} to features directly, we apply it to the predicted probabilities assigned to a set of samples by the model \(M\). If the model shifts, the probabilities assigned to each sample will change; again, this metric will equal 0 only if the two classifiers are the same. We compute PP MMD in two ways: firstly, we compute it over samples drawn uniformly from the dataset, and, secondly, we compute it over points spanning a mesh grid over a subspace of the entire feature space. For the latter approach, we bound the subspace by the extrema of each feature. While this approach is theoretically more robust, unfortunately, it suffers from the curse of dimensionality, since it becomes increasingly difficult to select enough points to overcome noise as the dimension \(D\) grows.

As an alternative to PP MMD, we use a pseudo-distance for the \textbf{Disagreement Coefficient} (Disagreement). This metric was introduced in \protect\hyperlink{ref-hanneke2007bound}{{[}28{]}} and estimates \(p(M(x) \neq M^\prime(x))\), that is the probability that two classifiers disagree on the predicted outcome for a randomly chosen sample. Thus, it is not relevant whether the classification is correct according to the ground truth, but only whether the sample lies on the same side of the two respective decision boundaries. In our context, this metric quantifies the overlap between the initial model (trained before the application of AR) and the updated model. A Disagreement Coefficient unequal to zero is indicative of a model shift. The opposite is not true: even if the Disagreement Coefficient is equal to zero, a model shift may still have occurred. This is one reason why PP MMD is our preferred metric.
